---
output:
  xaringan::moon_reader:
    #lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
    css: ["custom.css", default, shinobi, ninjutsu]
  pdf_document: default
  md_document: default
---

## Last week

- What is R?
- What is RStudio?
- Packages
- Variables and data types


---

## This week


--

- Data structures
  + Vectors
  + Lists
  + Dataframes
  
--

- Functions
  + What is a function?
  + How to use a function

---

## Data structures

- Values in R need to be stored in a specific way
- There are 4 main data structures in R:

---

## Data structures - Vectors

- Vectors

--

  + Vectors in R are arrays of data in one dimension

--

  + They are atomic (not recursive)*, but can have named values

--

  + Even variables with a single value will be stored as a vector with length 1

--


```{r data_structures_vectors, echo = TRUE}

vector_1 <- c(1,2,3,4,5,6)
vector_1

vector_2 <- c("first_value" = 1)
vector_2
```

- *More on the difference between atomic and recursive later
---

## Data structures - Data frames

- Data frames
  +  Most data you'll be working with in R will be held in a data frame
  
--
  
  +  Data frames are two dimensional arrays that can hold values of different types and have named columns and rows

--

```{r data_structures_dataframes, echo = TRUE}
dataframe1 <- data.frame(numeric_col = c(1,2),
                         character_col = c("hello", "world"),
                         stringsAsFactors = FALSE)
dataframe1
```

- Important notes:

--

  + Values in the same column must be of the same type

--

  + Each column must have the same number of rows

---

## Data structures - Lists

- Lists
  + Lists are similar to vectors in that they can contain multiple named values

--

  + However, lists can contain values of any type, including other lists

--

  + This means you can have a list of data frames, or a list of lists of data frames!
  
--

```{r data_structures_lists_1, echo = TRUE}

list_1 <- list(numbers = c(1,2,3,4),
               characters = (c("hello", "world"))
)
list_1

```

---

## Data structures - Lists

- Due to this difference between lists and vectors, we refer to a list as being *recursive* and vectors as *atomic*

--

- Atomic
  + "of or forming a single irreducible unit or component in a larger system."
    
--

  + Cannot hold objects of their own type (i.e. you can't have a vector object in a vector)
  
--
  
- Recursive
  + "characterized by recurrence or repetition."
  
--

  + Can hold objects of their own type (i.e you can have a list, in a list, in a list...)
  
---

## Lists vs vectors

``` {r, echo = TRUE}
vector_example <- c(1,2,3)
list_example <- list(first_vector = c(1,2,3),
                     first_list = list(1,2,3))
vector_example
list_example
```

---

## Data structures - Matrices

- Matrices

--

  + Matrices are two-dimensional arrays that hold values of the same type
  
--

  + Matrices can have named rows or columns, but they cannot be subsetted by those names (more later on)
  
--

```{r data_structures_matrices, echo = TRUE}

matrix_1 <- matrix(c(1,2, 3,4), nrow = 2, ncol = 2)
print(matrix_1)

```

---

## Subsetting data structures

- Most of the time you won't want all of the values in a data structure. For instance, usually you'll only want a column or two from a dataframe

--

- We call the process of extracting particular values from a data strucuture subsetting

--

- We're going to look very briefly at how to subset vectors, lists and dataframes, but this is covered in more detail in the book

--- 

## Subsetting lists and vectors


- To subset a list or vector, we use `[]` brackets

- In the brackets, we give the locations (indexes) of the values we want

--

Example:

```{r, echo = TRUE}

c(1,2,3)[1]
list(c(1,2,3), c(1,2,3))[1]
```

---

## Subsetting lists

- When subsetting lists, there is a difference between subsetting with `[]` and `[[]]`


--

- `[]`
  + Returns a list
  + Accepts more than one value

--

- `[[]]`
  + Returns a single object
  + Only accepts one value

---

## Example

```{r}
test_list <- list(first_value = c(1,2,3),
                  second_value = c(100,200,300))
test_list[1]
test_list[[2]]
```

---

## Subsetting lists and dataframes

- You can subset dataframes and lists (if they have named values) using `$`

--

```{r}
test_list$first_value
dataframe1$numeric_col
```
--

- For the list, this returns the item with that name

- For the dataframe, it returns the column

---

## Functions

- Any operation, such as loading a dataset or finding the mean, is performed via a function in R

--

- A function is a set of steps that takes an input, does some form of computation, and returns an output

--

- For example, the mean function takes values and returns the mean of those values

--

```{r functions_1, echo = TRUE}
values <- c(10,20,30)
mean(values)

```

--

- You may have noticed that `c()` also acts as a function, creating a vector with the provided values

---

## Functions - input parameters/arguments

- Almost all functions will require some input to produce an output

--

- These are the function's input parameters or arguments

--

- For example, in the previous slide, the mean() function required a vector of values to calculate the mean

---

## Functions - input parameters/arguments

- Some functions will have optional input parameters

--

- To see what arguments/parameters a function requires, type the function name preceded by a "?" in the R console (e.g. `?mean`)

---

## Functions - exercise

- Find the help page for the `sample()` function

--

- Use the `sample()` function

---

## Functions - naming input parameters

- When you provide unnamed values to a function, R will automatically assign them to an input argument in the order they appear on the function's help page

--

- This can be hard to keep track of if you're providing multiple arguments

--

- Best practice, therefore, is to be explicit when providing your input arguments

--

- Let's use the `substr()` function as an example

---

## Functions - naming input parameters


- The `substr()` function returns part of a string from a start and stop index

--

```{r function_naming_input_parameters_1, echo = TRUE}
string1 <- "hello world"
substr(string1, 1,5)
```

- The `substr()` function expects 3 arguments;

--

- The string (x)

--

- The start point (start)

--

- The end point (stop)

--

- As shown above, we can just pass those arguments without naming them, in that order

---

## Functions - naming input parameters

- It's best practice, however, to name the arguments as you pass them...

--

```{r function_naming_input_parameters_2, echo = TRUE}
string1 <- "hello world"
substr(x = string1, start = 1, stop = 5)
```

---

## Recap

- Values can then be stored in lots of different structures*
- These structures can be subsetted using different approaches
- Almost every action in R is performed via a function*
- A function takes an input, does some computation and returns an output
- It's always good to be explicit when providing arguments to functions
