---
title: "Data Analysis (Part 2)"
output:
  xaringan::moon_reader:
    #lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
    css: ["custom.css", default, shinobi, ninjutsu]
  pdf_document: default
  md_document: default
---

```{r, echo = FALSE}
library(readr)
c19 <- read_csv("../../data/C19.csv", col_types = cols(col_date(format = "%d/%m/%Y"),
                                                       col_date(format = "%d/%m/%Y"),
                                                       col_date(format = "%d/%m/%Y"),
                                                       col_character(),
                                                       col_factor(),
                                                       col_character(),
                                                       col_factor(),
                                                       col_logical(),
                                                       col_logical(),
                                                       col_logical(),
                                                       col_logical()))
```

## Last week

- Tidyverse

--

- `%>%`

--

- Loading data

--

- Cleaning data

---

## This week

- Summarising data with `dplyr`

--

- Plotting with `ggplot2`

## Setup

- If you haven't already, install `ggplot2` and `dplyr` using the `install.packages()` function

- Add the `library(dplyr)` and `library(ggplot2)` lines to the top of your script (if they aren't already there)

- Make sure you've got your C19 dataset loaded

- Filter and sort your dataset by date and by age group with:

```{r}
library(dplyr)
c19 <- c19 %>%
  filter(cdc_report_dt >= as.Date("2020/03/01")) %>%
  arrange(desc(cdc_report_dt), age_group)
  
```

## Summarising data

- Our dataset currently has nearly 33,000 rows

--

- This is a lot of data, and so extracting meaningful conclusions from this level can be difficult

--

- Instead of analysing data at this level, you might want to aggregate it up

- Then you can always investigate further at a more granular level in the future but this will point you in the right direction

## Summarising data

- Using our C19 data, we're going to aggregate up to a daily level and for each age group, so we'll have a value for each combination of date and age group*

--

- \* Unless there are no rows for a date and age group (e.g. there are no entries for "20-29 Years" on Feb 2nd)

--

- This will take us down from ~33,000 rows to 235

--

- In `dplyr`, we use the `group_by()` and `summarise()` functions to aggregate our data up

--

- If you used SQL before, it's a similar approach

---

## `group_by` & `summarise()`

- Before we use the `summarise()` function, we need to tell R what we want variable we want to group by.

--

- This will then create a value for each unique value of that variable(s) that we grouped by

--

- So in our example, we would have a summarised value for each date and age group if we grouped by the `cdc_report_dt` and `age_group` columns


---

## `group_by`

- So we know that we want to group our data by the `cdc_report_dt` and `age_group`

--

- To do so, we use the `group_by()` function like so:

```{r}
library(dplyr, quietly = TRUE)
c19_grouped <- c19 %>%
  group_by(cdc_report_dt, age_group)
head(c19_grouped, 5)
```

--

- Now when we view our dataset, it looks like the same but we can see that it is grouped: `Groups: cdc_report_dt, age_group [3]`

## `sumamrise()`

- The data still looks the same though because we haven't done our summarisation yet, we've just told R what we want to group by

--

- To summarise our data, we use the `summarise()` function a bit like we used the `mutate()` function:

```{r}
c19_grouped <- c19 %>%
  group_by(cdc_report_dt, age_group) %>%
  summarise(number_of_cases = n()) %>%
  ungroup()
head(c19_grouped, 5)

```

---

## `summarise()`

- In the `summarise()` function call, we say what we want our new column's name to be (number_of_cases here) and then we need to say what kind of summary function we want to apply

--

- In this case, we've just counted the number of entries for that date and age group (that's what the `n()` function does)

--

- But we could use any other summary function

- For example, if we wanted to find the percentage of cases that resulted in death, we could use the `mean()` function on the `death_yn` column*:

```{r}
c19_example <- c19 %>%
  group_by(cdc_report_dt, age_group) %>%
  summarise(mortality_rate = mean(death_yn) * 100) %>%
  ungroup()
head(c19_example, 5)
```

--

- \*Because logical values `TRUE`/`FALSE` are equivalent to numeric `1` and `0`, we can use numeric functions on logical value

---

## Plotting

- Now that we have our summarised dataset, we want to actually see what's going on

--

- The easiest way to do this would be to create a line plot the shows the number of cases by each age group and date 

- This is quite a common graphic that's displayed during the Government's press briefings

--

- To create our plot, we're going to use the `ggplot2` package

---

## `ggplot2`

- The `ggplot2` package is a plotting library based on The Grammer of Graphics

--

- This is basically the idea that graphics are a bit like language in that they have a grammer that should be followed to make them understandable

--

- As part of these grammatical rules, we build plots in `ggplot2` in stages. You define your dataset, then what geom you want (lines, points, etc.), what you want to group the lines by, then what you want the scales to look like, etc.

--

- But rather than using the pipe, we use the `+` operator in `ggplot2`

--

- `ggplot2` is a very extensive and useful library, and we're really only going to brush the surface today

## `ggplot2`

- To start of with, we need to define what dataset we want to plot, what we want on each axis and what we want to group by

- We do this using the `ggplot()` function to define our dataset, alongside the `aes()` function to define our axes:

```{r}
library(ggplot2)
ggplot(data = c19_grouped, aes(x = cdc_report_dt, y = number_of_cases))
```

---

## `ggplot2`

- You'll notice that our plot is blank though

--

- This because we need to define what type of geom we want to use (e.g. bars, lines, points, etc.)

--

- In our case, we want to use a line, so we add the `geom_line()` function:

```{r}
ggplot(data = c19_grouped, aes(x = cdc_report_dt, y = number_of_cases)) +
  geom_line()
```

---

## `ggplot2`

- Better, but that's not really what we wanted...

--

- We're seeing these weird squigly lines because our dataset has multiple values for each date (we've got a value for each age group)

--

- So what we need to do is have separate lines for each of our age groups, and perhaps we should have those as different colours

--

- To do that, we need to change our original `aes()` call:

```{r}
ggplot(data = c19_grouped, aes(x = cdc_report_dt, y = number_of_cases, colour = age_group)) +
  geom_line()
```

---

## `ggplot`

- Much better

--

- But things are still messy...

- For instance, our axes labels are our variable names which don't look good. And also we probably don't want the "Unknown" and "NA" groups

--

- To change our scales (age_group also counts as a scale), we use the `scale_x/y_[type]` functions. For example, if we want to change the x axis and the x variable is a date, we'
d use the `scale_x_date()` function:

```{r}
ggplot(data = c19_grouped, aes(x = cdc_report_dt, y = number_of_cases, colour = age_group)) +
  geom_line() +
  scale_x_date(name = "CDC Report Date") +
  scale_y_continuous(name = "Number of Cases") +
  scale_color_discrete(name = "Age Group")
```

---

## Exercises

- Remove the "Unknown" and "NA" groups from your dataset using the `filter()` function and recreate your plot

- Change the date breaks so that there is a label every two months using the `date_breaks` parameter of the `scale_x_date()` function (use `?scale_x_date()` to find out how to use it)

- Change your dataset and plot so that you're plotting the mortality rate instead of the number of cases
